---
title: "432 Class 03"
author: "<https://thomaselove.github.io/432-2023/>"
date: "2023-01-24"
date-format: iso
format: 
  beamer:
    theme: Madrid
    colortheme: lily
    fig-align: center
---

## Today's Agenda

Incorporating Survey Weights ...

1. In estimating means and confidence intervals
2. In building linear regression models
3. Into a more detailed t-test approach using NHANES

Primary Source: <https://bookdown.org/rwnahhas/RMPH/survey-design.html>

## Today's R Setup

```{r}
#| echo: true
knitr::opts_chunk$set(comment = NA)

library(broom)
library(janitor) 
library(gtsummary)
library(knitr)
library(mosaic)

library(nhanesA) # data source
library(survey) # survey-specific tools

library(tidyverse)

theme_set(theme_bw())
```

# Incorporating survey weights (an introduction)

## What are survey weights?

In many surveys, each sampled subject is assigned a weight that is equivalent to the reciprocal of his/her probability of selection into the sample.

$$
\mbox{Sample Subject's Weight} = \frac{1}{Prob(selection)}
$$

but more sophisticated sampling designs require more complex weighting schemes. Usually these are published as part of the survey data.

There are several packages available to help incorporate survey weights in R, but I will demonstrate part of the `survey` package today.

## An NHANES Example

Let's use the NHANES 2013-14 data and pull in both the demographics (`DEMO_H`) and total cholesterol (`TCHOL_H`) databases.

```{r}
#| echo: true
demo_raw <- nhanes('DEMO_H')
tchol_raw <- nhanes('TCHOL_H')
```

Detailed descriptions available at

- <https://wwwn.cdc.gov/Nchs/Nhanes/2013-2014/DEMO_H.htm>
- <https://wwwn.cdc.gov/Nchs/Nhanes/2013-2014/TCHOL_H.htm>

## Weighting in NHANES

Weights are created in NHANES to account for the complex survey design. A sample weight is assigned to each sample person. It is a measure of the number of people in the population represented by that sample person.

The sample weight is created in three steps:

1. the base weight is computed, which accounts for the unequal probabilities of selection given that some demographic groups were over-sampled;
2. adjustments are made for non-response; and
3. post-stratification adjustments are made to match estimates of the U.S. civilian non-institutionalized population available from the Census Bureau.

Source: <https://wwwn.cdc.gov/nchs/nhanes/tutorials/Module3.aspx>


## Weights in our NHANES data

The `DEMO` file contains two kinds of sampling weights:

- the interview weight (`WTINT2yr`), and
- the MEC exam weight (`WTMEC2yr`)

NHANES also provides several weights for subsamples. A good rule for NHANES is to identify the variable of interest that was collected on the smallest number of respondents. The sample weight that applies to that variable is the appropriate one to use in your analysis.

In our first case, we will study total cholesterol and use the weights from the MEC exam.

## What Variables Do We Need?

- `SEQN` = subject identifying code
- `RIAGENDR` = sex (1 = M, 2 = F)
- `RIDAGEYR` = age (in years at screening, topcode at 80)
- `DMQMILIZ` = served active duty in US Armed Forces (1 = yes, 2 = no)
- `RIDSTATR` = 2 if subject took both interview and MEC exam
- `WTMEC2YR` - Full sample 2 year MEC exam weight
- `LBXTC` = Total Cholesterol (mg/dl) - this is our outcome

The first five of these came from the `DEMO_H` file, and the first and last comes from `TCHOL_H`.

## Merge the `DEMO` and `TCHOL` files

```{r}
#| echo: true

dim(demo_raw)
dim(tchol_raw)

joined_df <- inner_join(demo_raw, tchol_raw, by = c("SEQN"))

dim(joined_df)
```

## Create a small analytic tibble

```{r}
#| echo: true

nh1314 <- joined_df |> # has n = 8291
    tibble() |>
    filter(complete.cases(LBXTC)) |> # now n = 7624
    filter(RIDSTATR == 2) |> # still 7624
    filter(RIDAGEYR > 19 & RIDAGEYR < 40) |> # now n = 1802
    filter(DMQMILIZ < 3) |> # drop 7 = refused, n = 1801
    mutate(FEMALE = RIAGENDR - 1,
           AGE = RIDAGEYR,
           US_MIL = ifelse(DMQMILIZ == 1, 1, 0),
           WT_EX = WTMEC2YR,
           TOTCHOL = LBXTC) |>
    select(SEQN, FEMALE, AGE, TOTCHOL, US_MIL, WT_EX)
```

## Our `nh1314` analytic sample: Variables

```{r}
#| echo: true
nh1314 |> tabyl(FEMALE, US_MIL) |> 
  adorn_totals(where = c("row", "col")) |> adorn_title()
```

```{r}
#| echo: true
df_stats(~ AGE + TOTCHOL, data = nh1314) |>
  rename(med = median, na = missing) |>
  kable(digits = 1)
```

## Our `nh1314` analytic sample: Weights

Each weight represents the number of people exemplified by that subject.

```{r}
#| echo: true
favstats(~ WT_EX, data = nh1314) |>
  rename(na = missing) |>
  kable(digits = 1)
```

## Create `nh_design` survey design

```{r}
#| echo: true
nh_design <- 
    svydesign(
        id = ~ SEQN,
        weights = ~ WT_EX,
        data = nh1314) 

nh_design <- update( nh_design, one = 1) 

## this one = 1 business will help us count
```

## Unweighted counts, overall and by sex

```{r}
#| echo: true

sum(weights(nh_design, "sampling") != 0)
svyby( ~ one, ~ FEMALE, nh_design, unwtd.count)
svyby( ~ one, ~ FEMALE + US_MIL, nh_design, unwtd.count)
```

## Weighted counts, overall and by groups

Weighted size of the generalizable population, overall and by groups.

```{r}
#| echo: true

svytotal( ~ one, nh_design )

svyby( ~ one, ~ FEMALE * US_MIL, nh_design, svytotal)
```

## Use the survey design to get weighted means

What is the mean of total cholesterol, overall and in groups?

```{r}
#| echo: true

svymean( ~ TOTCHOL, nh_design, na.rm = TRUE)

svyby(~ TOTCHOL, ~ FEMALE + US_MIL, nh_design, 
      svymean, na.rm = TRUE)
```

## Unweighted Mean of TOTCHOL

```{r}
#| echo: true

nh1314 |> 
  summarise(n = n(), mean(TOTCHOL)) |>
  kable(digits = 2)
```

Note that we're using `summarise` to ensure that we get the **dplyr** package's version of `summarize`.

## Unweighted Group Means of TOTCHOL

```{r}
#| echo: true

nh1314 |> group_by(FEMALE, US_MIL) |>
  summarise(n = n(), mean(TOTCHOL)) |>
  kable(digits = 2)
```


## Measures of uncertainty (Survey-Weighted)

Mean of total cholesterol within groups with 90% CI?

```{r}
#| echo: true

grouped_result <- svyby(~ TOTCHOL, ~ FEMALE + US_MIL, 
                        nh_design, svymean, na.rm = TRUE)
coef(grouped_result)
confint(grouped_result, level = 0.90)
```

- Get standard errors with `se(grouped_result)`, too.

## Placing estimated means in `res`

```{r}
#| echo: true
res <- tibble(
  type = rep(c("Unweighted", "Survey-Weighted"),4),
  female = c(rep("Female", 4), rep("Male", 4)),
  us_mil = rep(c("Military", "Military", 
                 "Not Military", "Not Military"), 2),
  MEAN = c(169.5, 164.1984, 179.71, 180.0248, 
           187.11, 186.6966, 182.22, 182.3569) )
```

## The Estimated Means

```{r}
#| echo: true
res |> kable(digits = 1)
```

```{r}
#| echo: true
#| eval: false
ggplot(res, aes(x = female, y = MEAN, col = type)) +
  geom_point(size = 4) +
  facet_wrap(~ us_mil) ## plot shown on next slide
```

## Plotting the Estimated Means

```{r}
ggplot(res, aes(x = female, y = MEAN, col = type)) +
  geom_point(size = 4) +
  facet_wrap(~ us_mil)
```

# Building Models

## Models for `TOTCHOL` in our `nh1314` data

First, we'll ignore the weighting, and fit one model with main effects of all three predictors (model mod1) and then a second model which incorporates an interaction of FEMALE and US_MIL.

```{r}
#| echo: true
mod1 <- lm(TOTCHOL ~ AGE + FEMALE + US_MIL, data = nh1314)

mod2 <- lm(TOTCHOL ~ AGE + FEMALE * US_MIL, data = nh1314)
```

The interaction term means that the effect of FEMALE on TOTCHOL depends on the US_MIL status.

## Unweighted Model `mod1` (no interaction)

```{r}
#| echo: true
tidy(mod1, conf.int = TRUE, conf.level = 0.90) |>
  select(-statistic) |> kable(digits = 2)
```

```{r}
#| echo: true
glance(mod1) |> select(r2 = r.squared, adjr2 = adj.r.squared,
    AIC, BIC, sigma, nobs, df) |> kable(dig = c(4,4,1,1,3,0,0))
```

## Residuals for Model `mod1`

```{r}
par(mfrow = c(2,2)); plot(mod1)
```

## Unweighted Model `mod2` (with interaction)

```{r}
tidy(mod2, conf.int = TRUE, conf.level = 0.90) |>
  select(-statistic) |> kable(digits = 2)
```

```{r}
glance(mod2) |> select(r2 = r.squared, adjr2 = adj.r.squared,
    AIC, BIC, sigma, nobs, df) |> kable(digits = c(4,4,1,1,3,0,0))
```


## Residuals for Model `mod2`

```{r}
par(mfrow = c(2,2)); plot(mod2)
```

## Perform a survey-weighted generalized linear model

Again, we'll run two models, first without and second with an interaction term between `FEMALE` and `US_MIL`.

```{r}
#| echo: true

glm1_res <- svyglm(
    TOTCHOL ~ AGE + FEMALE + US_MIL, 
    nh_design, family = gaussian())
```

```{r}
#| echo: true

glm2_res <- svyglm(
    TOTCHOL ~ AGE + FEMALE * US_MIL, 
    nh_design, family = gaussian())
```

Gaussian family used to generate linear regressions here.

## Model 1 Results

```{r}
#| echo: true

tidy(glm1_res, conf.int = TRUE, conf.level = 0.90) |>
  select(-statistic) |> kable(digits = 2)
```

```{r}
#| echo: true

glance(glm1_res) |>
  select(nobs, AIC, BIC, everything()) |> kable(digits = 1)
```

## Model 2 Results

```{r}
#| echo: true

tidy(glm2_res, conf.int = TRUE, conf.level = 0.90) |>
  select(-statistic) |> kable(digits = 2)
```

```{r}
#| echo: true

glance(glm2_res) |>
  select(nobs, AIC, BIC, everything()) |> kable(digits = 1)
```

## Residuals for Model `glm1_res`

```{r}
par(mfrow = c(2,2)); plot(glm1_res)
```

## Residuals for Model `glm2_res`

```{r}
par(mfrow = c(2,2)); plot(glm2_res)
```

# A More Complete Weighted NHANES Analysis (See <https://wwwn.cdc.gov/nchs/data/tutorials/DB303_Fig1_R.R>)

## New Question, New Data

We are now interested in looking at the percentage of persons aged 20 and over with depression, by age and sex, in the US in 2013-2016. Start by pulling in the relevant data using `nhanesA`...

```{r}
#| echo: true

DEMO_H <- nhanes('DEMO_H') |> select(SEQN, RIAGENDR, 
                    RIDAGEYR, SDMVSTRA, SDMVPSU, WTMEC2YR)
DEMO_I <- nhanes('DEMO_I') |> select(SEQN, RIAGENDR, 
                    RIDAGEYR, SDMVSTRA, SDMVPSU, WTMEC2YR)
DEMO <- bind_rows(DEMO_H, DEMO_I) 
DPQ_H <- nhanes('DPQ_H') 
DPQ_I <- nhanes('DPQ_I') 
DPQ <- bind_rows(DPQ_H, DPQ_I) 
```

## Merge DEMO and DPQ files and create derived variables

```{r}
#| echo: true
dat2 <- left_join(DEMO, DPQ, by = "SEQN") |> tibble() |>
  # Set 7=Refused and 9=Don't Know To NA 
  mutate(across(.cols = DPQ010:DPQ090, 
                ~ ifelse(. >=7, NA, .))) %>%
  mutate(one = 1,
         PHQ9_score = rowSums(select(. , DPQ010:DPQ090)),
         Depression = ifelse(PHQ9_score >= 10, 100, 0),
         Sex = factor(RIAGENDR, labels = c("M", "F")),
         Age_group = cut(RIDAGEYR, 
            breaks = c(-Inf, 19, 39, 59, Inf),
            labels = c("Under 20", "20-39", "40-59", "60+")),
         WTMEC4YR = WTMEC2YR/2,
         inAnalysis = (RIDAGEYR >= 20 & !is.na(PHQ9_score))) |>
  select(-starts_with("DPQ"))
```

## Define Survey Design

Here's the survey design for the overall data set:

```{r}
#| echo: true
NH_des_all <- svydesign(data = dat2, id = ~ SDMVPSU, 
  strata = ~ SDMVSTRA, weights = ~ WTMEC4YR, nest = TRUE)

dim(NH_des_all)
```

Here's the survey design object for the subset of interest: adults aged 20 and over with a valid PHQ-9 depression score:

```{r}
#| echo: true

NH_des_dat2 <- NH_des_all |> subset(inAnalysis)

dim(NH_des_dat2)
```

## Define a function to call svymean and unweighted count

```{r}
#| echo: true

ourSummary <- function(varformula, byformula, design){
  # Get mean, stderr, and unweighted sample size
  c <- svyby(varformula, byformula, design, unwtd.count ) 
  p <- svyby(varformula, byformula, design, svymean ) 
  outSum <- left_join(select(c,-se), p) 
  outSum
}
```

### Estimate overall prevalence of depression

```{r}
#| echo: true
ourSummary(~ Depression, ~ one, NH_des_dat2)
```

## Estimate prevalence of depression in various strata

```{r}
#| echo: true

## By sex
ourSummary(~ Depression, ~ Sex, NH_des_dat2)
## By age
ourSummary(~ Depression, ~ Age_group, NH_des_dat2)
```

## Estimate prevalence of depression by Age and Sex

```{r}
#| echo: true

## By sex and age
ourSummary(~ Depression, ~ Sex + Age_group, NH_des_dat2)
```

## Compare Prevalence between Male and Female

Across all age groups:

```{r}
#| echo: true
svyttest(Depression ~ Sex, NH_des_dat2)
```

## Compare Prevalence between Male and Female

In people ages 40-59:

```{r}
#| echo: true
svyttest(Depression ~ Sex, subset(NH_des_dat2, Age_group == "40-59"))
```

## Differences by Age Group, among Adults

```{r}
#| echo: true

svyttest(Depression ~ Age_group, subset(NH_des_dat2, 
                  Age_group=="20-39" | Age_group=="40-59"))
```

## Next Time?

Linear Regression and ANOVA / ANCOVA models

### Reminders

1. Please complete the **Minute Paper after Class 3** by noon tomorrow (Wednesday 2023-01-25)
2. Get started on **Lab 2**, due next Monday 2023-01-30 at 9 PM
3. Continue reading **How To Be A Modern Scientist**
